INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-11-06 chapters/background.tex
addressing comments

\PV{Nasledujici dve vety byly vzaty z popisu INGRID, tak funguje INGRID, ne PE4MPS.}
An MPS language is created by the PE4MPS tool in two steps.
First, the PE parser is used to load an input PE file and build its intermediate representation, which has the form of a tree-like structure consisting of Java objects.
Then, all the concepts (AST nodes) and their aspects are created inside MPS.
The information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.

The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
\PV{Navrhuji pridat (aby to bylo vice jasne?): User is forced to input the layout information by extending the grammar manually.}
Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

@@

However, the project has also other important limitations.
The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
\PV{Dalsi veta mozna zbytecna?}
In particular, it does not support creating of the projectional editor at all.

\PV{Navrhuji pridat neco jako (ale mozna uz o tom mluvime jinde, jen si ted nejsem jisty):
The major take away from this project is the idea of an intermediate grammar representation inside of MPS which we have also considered.
The advantage of this could be the possibility of extending the ANTLRv4 notation using some graphic tools that MPS has to offer. 
Using the projectional editor, a user could add the layout information in an effective way.
The disadvantage is the dual implementation that is needed.
First to transform the ANTLRv4 text file into the MPS grammar language and secondly to transform this grammar into the target MPS language.}

\paragraph{mps-metabnf.}
The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
Authors of this project, too, created an MPS language that can be used to capture ANTLR grammars.
Every imported grammar is represented using this ANTLRv4 MPS language.
A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
In the last step, the target MPS language is derived from the adjusted grammar definition.
\PV{Navrhuji odsud dolu smazat.}
Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.

The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.
\PV{Smazat az sem.}
\PV{A pridat: This step has not yet been implemented, because MPS doesn't offer any mechanisms for automatic transformation.
This means that implementation would require traversing the AST of the grammar and recreating the definition from that.}

>>>>

The PE4MPS tool creates an MPS language in a single atomic step.
An input PE file is loaded and processed by the PE parser, and then all concepts (AST nodes) and their aspects are directly generated inside MPS.
Information about the code layout, extracted from the PE file, is used when generating the projectional editor for every AST node.

The main disadvantage of the PE4MPS approach is that it only shifts the tedious manual work from creating projectional editors in the MPS IDE to writing PE grammars in plain text.
In particular, language developers are forced to specify the code layout manually in PE grammars, through the corresponding extensions.
Usage of a standard text editor is inferior (and much more error-prone) with respect to MPS, which has been designed for the purpose of specifying the code layout and provides lot of support to developers.
Another limitation of PE4MPS is that it does not implement the full ANTLR syntax.
Every grammar might require non-trivial adjustments before it can be processed using the PE4MPS tool.
To compare, the goals behind our INGRID approach are (1) to enable import of as many languages as possible out-of-the-box with maximal automation and (2) to adopt the full specification of every imported language.

@@

However, the project has also other important limitations.
The tool does not generate any parent-child relationships in the Structure aspect, and it provides no support for the projectional editor at all (by completely neglecting the Editor and TextGen aspects).

\paragraph{mps-metabnf.}
The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
Authors of this project, too, created an MPS language that can be used to capture ANTLR grammars.
Every imported grammar is represented using this ANTLRv4 MPS language.
A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
In the last step, the target MPS language is derived from the adjusted grammar definition.

The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the definition of the actual target MPS language.

### 2016-11-06 chapters/language_definition.tex
evaluation - first part
evaluation - done
addressing comments

The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.
\PV{Mozna vysvetlit proc (protoze je to dvoji prace) nebo odkazat tam, kde je to vysvetleny?}

@@

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.
An output of this phase is a custom AST-like structure representing the grammar.

\PV{Tenhle dalsi odstavec bych klidne celej odebral, neprinasi zas tolik informaci, vzhledem k tomu, ze nejde do hloubky.}
A simplified tree of objects is derived in two passes over the grammar.
The result of the first pass is a tree where each node contains names of other referenced rules in the form of strings.
In the second pass, the whole tree is traversed and all the references are resolved to have a form of real pointers to other rule objects.
\PV{Az sem}

@@

\PV{Tady bych na zacatek vety pridal neco jako "In MPS...". Neni jasne, jestli mluvime o necem v ingrid nebo o konceptech MPS.}
Each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
In addition, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar (i.e., from the structure of the AST).
For each rule, the object corresponding to the left-hand side is in the parent-child relationships with sets of objects that represent language elements referenced by the right-hand side.
If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created in the structure for each alternative.
The name of a concept (object) in MPS is composed from (1) the name of the AST node, which is equivalent to the string encoding of the left-hand side of the corresponding rule, and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.

@@

Consider the parser rule \antlrparserrule{element} from the grammar in Figure~\ref{fig:SIMPLEXML}.
Its first alternative represents the full XML element with content.
Figure~\ref{fig:ELEMENTSTRUCT} shows the Structure aspect for the MPS concept (object) named \code{Element{\_}1} that corresponds to the alternative.
The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
\PV{Dalsi vetu bych nahradil: }
\PV{Values of these properties will be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} lexer rule.}
\PV{Protoze }
Values of these properties should be restricted using the regular expression that corresponds to the \antlrlexerrule{Name} rule, but this is left to the human user.
\PV{Mozna "for" nahradit necim z XML, treba zavorkami </>, najednou totiz se preskoci na for, ktery se vzal odnikud.}
String literals, such as the keyword \textbf{for}, are omitted because they will be defined only in the Editor aspect for this concept.

References to other parser rules are captured by pointers to child objects.
The types of child objects, such as \code{IContent} in the \code{Element{\_}1} concept, are determined as follows.
Consider the \antlrparserrule{content} rule from the SimpleXML language.
\PV{Label CONTENTRULE defined multiple times! Pak to ukazuje spatny cislo Figure}
We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
An object corresponding to any one of the four alternatives could be the actual value anywhere the \antlrparserrule{content} rule is referenced.

@@

Our solution is to use interface concepts.
\PV{Odstranil bych ", i.e. for each AST node with more than one child"}
For each rule with more than one alternative on the right side, i.e. for each AST node with more than one child, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
The resulting fragment of the language structure looks like the one in Figure~\ref{fig:ICONTENTITF}.
It contains one interface \mpsinterface{IContent} that is implemented by four object concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.

@@

The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTRULE} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
  \antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
  \antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
  \antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\end{alltt}
\end{framed}
\caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
\label{fig:CONTENTRULE}
\end{figure*}

@@

We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTRULE}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.

@@

\subsection{Phase 3: Editor}

@@

\subsection{Phase 4: Text Generation}

@@

\PV{Nejsem si jisty, jestli to nebylo az pozdeji, ale byl tu nekde odstavec, ktery se zminoval o tom, ze whitespace je silent-polykan parserem a proto pravidla nemusi obsahovat explicitne informaci o tom, kde vsude muze byt mezera \\\texttt{'<' WS* Name WS* attribute* WS* '>' WS*...}.\\To podle me dobre ilustrovalo, ze se tam ztraci ta informace. Prislo mi to jako zajimave poukazani, ale samozrejme to tu byt nemusi, pokud se ti zda, ze to az tak zajimave neni.}

@@

The second example relates to a fragment of the JavaScript language, also known as ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.
Since there are multiple options, a user would have to select one of them for each statement in the editor.
Technically, every language concept representing a statement would contain one child node of the interface type \mpsinterface{IEos}, which has to be assigned a proper object (i.e., an AST node corresponding to one of the options listed above).

>>>>

The proposed INGRID method accepts grammars in the ANTLR v4 notation~\cite{ref:ANTLRBOOK} as input.
All widely used programming languages have their syntax defined using this notation~\footnote{https://github.com/antlr/grammars-v4}.
Unlike some of the related projects, we did not extend the notation with any custom features, and we also did not create an MPS language for the ANTLR notation.

@@

The first phase in the process of MPS language construction is parsing of the input ANTLR v4 grammar.
It is done using an ANTLR parser that was automatically generated from the grammar of the ANTLR v4 notation itself.
Nevertheless, the full AST, which comes out of the parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy to process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST are performed in this phase.
An output is a custom AST-like structure representing the grammar, especially the hierarchy of parser rules.

@@

In MPS, each concept of the language (i.e., every AST node) is represented by an object that may have a parent, some children, and properties of any data type.
Additionally, the object may implement any number of interfaces, and it may also contain references to objects representing other AST nodes.
The parent-child relationships between objects that make the Structure aspect are derived from rules of the grammar (i.e., from the structure of the AST).
For each rule, the object corresponding to the left-hand side is in the parent-child relationships with sets of objects that represent language elements referenced by the right-hand side.
If a rule has multiple alternatives, then a distinct object (MPS concept) has to be created in the structure for each alternative.
The name of a concept (object) in MPS is composed from (1) the name of the AST node, which is equivalent to the string encoding of the left-hand side of the corresponding rule, and (2) the number indicating the position of the respective alternative on the right-hand side of the rule.

@@

Consider the parser rule \antlrparserrule{element} from the grammar in Figure~\ref{fig:SIMPLEXML}.
Its first alternative represents the full XML element with content.
Figure~\ref{fig:ELEMENTSTRUCT} shows the Structure aspect for the MPS concept (object) named \code{Element{\_}1} that corresponds to the alternative.
The object contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name}.
Values of these properties will be restricted using the regular expression that corresponds to the lexer rule \antlrlexerrule{Name}.
String literals, such as the opening and closing brackets (\antlrliteral{\textless}, \antlrliteral{/\textgreater}) in XML, are omitted because they will be defined only in the Editor aspect for this concept.

References to other parser rules are captured by pointers to child objects.
The types of child objects, such as \code{IContent} in the \code{Element{\_}1} concept, are determined as follows.
Consider the \antlrparserrule{content} rule from the SimpleXML language.
We show just the relevant fragment of the grammar again in Figure~\ref{fig:CONTENTRULE}.
An object corresponding to any one of the four alternatives could be the actual value anywhere the \antlrparserrule{content} rule is referenced.

@@

Our solution is to use interface concepts.
For each rule with more than one alternative on the right side, first an interface concept is defined in the MPS language structure, and then one object that implements the given interface is created for each alternative.
The resulting fragment of the language structure looks like the one in Figure~\ref{fig:ICONTENTITF}.
It contains one interface \mpsinterface{IContent} that is implemented by four object concepts \mpsconcept{Content{\_}1}, $\ldots$, \mpsconcept{Content{\_}4}.

@@

The key idea of this approach is to skip all the intermediary layers (nodes) in the syntax tree and consider just the leaf nodes, e.g., to be directly offered to the user through the auto-completion menu.
Specifically, the \antlrparserrule{content} rule from the SimpleXML grammar, which is given in Figure~\ref{fig:CONTENTHIERARCHY} together with rules that determine the relevant fragment of the syntax hierarchy, expands ultimately into the leaf nodes that are highlighted using the bold font in Figure~\ref{fig:CONTENTEXPAND}.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
  \antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
  \antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
  \antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\end{alltt}
\end{framed}
\caption{The parser rule \antlrparserrule{content} with other rules that determine the relevant fragment of the syntax hierarchy}
\label{fig:CONTENTHIERARCHY}
\end{figure*}

@@

We use the name \emph{shortcut approach} for this algorithm, because the paths collected by the algorithm provide shortcuts from the given rule to end nodes, by the virtue of hiding all intermediary layers.
For example, the result of this algorithm for the \antlrparserrule{content} rule (Figure~\ref{fig:CONTENTHIERARCHY}) is the list of five items already shown in Figure~\ref{fig:CONTENTEXPAND}.

@@

\subsection{Phase 3: Editor}
\label{sect:EDITORDEF}

@@

\subsection{Phase 4: Text Generation}
\label{sect:TEXTGENDEF}

@@

@@

The second example relates to a fragment of the JavaScript language, also known as ECMAScript.
%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.
Since there are multiple options, a user would have to select one of them for each statement in the editor.
Technically, every language concept representing a statement would contain one child node of the interface type \mpsinterface{IEos}, which has to be assigned a proper object (i.e., an AST node corresponding to one of the options listed above).
